\section{Error Analysis}

In this section, we establish an error estimation framework for the proposed Indirect PINN method with hard constraint. We decompose the total error into several components, derive a conditional a priori error bound, and construct a computable a posteriori error indicator that is consistent with the meshfree, collocation-based nature of PINNs. Unlike finite-element residual estimators, our indicator does not rely on a computational mesh and therefore contains no inter-element jump terms.

\subsection{The Optimality System and Its Residuals}

Recall the first-order optimality system for our problem: find $(y, u, \lambda)$ such that

\begin{align}
\partial_t y - \nu \Delta y + f(y)-g &= 0, \quad &&\text{in } Q, \label{eq:state_res}\\
-\partial_t \lambda - \nu \Delta \lambda + f'(y) \lambda +(y-y_d)&=0, \quad &&\text{in } Q, \label{eq:adjoint_res}\\
\alpha u +\nu \partial_n \lambda &= 0, \quad &&\text{on } \Sigma, \label{eq:stat_res}\\
y &= u, \quad &&\text{on } \Sigma, \label{eq:bc_y_res}\\
\lambda &= 0, \quad &&\text{on } \Sigma, \label{eq:bc_lambda_res}\\
y(x,0) &= y_0(x), \quad &&\text{in } \Omega, \\
\lambda(x,T) &= 0, \quad &&\text{in } \Omega.
\end{align}

This is precisely the system \eqref{eq:state_opt}--\eqref{eq:bc_lambda_res} of Section~2, and we use the same constant $\alpha$ as in the cost functional \eqref{eq:cost} (we no longer introduce the separate symbols $\beta_Q,\beta_R$ used in an earlier draft, to keep notation consistent across sections; the tracking term always carries weight $1$ and the control term always carries weight $\alpha$).

Let $Y_\theta$, $U_\psi$, $\Lambda_\phi$ be the neural network approximations built with the hard constraints of Section~2.2.2. The state and adjoint boundary conditions and the adjoint terminal condition hold exactly, whereas the state initial condition is penalized. We define

\begin{align}
\mathcal{R}_y(x,t;\theta,\psi) &:= \partial_t Y_\theta - \nu \Delta Y_\theta + f(Y_\theta)-g, \label{eq:res_y}\\
\mathcal{R}_\lambda(x,t;\theta,\phi) &:= -\partial_t \Lambda_\phi - \nu \Delta \Lambda_\phi + f'(Y_\theta) \Lambda_\phi + (Y_\theta - y_d), \label{eq:res_lambda}\\
\mathcal{R}_u(x,t;\psi,\phi) &:= \alpha U_\psi +\nu\partial_n \Lambda_\phi, \quad (x,t) \in \Sigma. \label{eq:res_u}
\end{align}

Because the boundary and adjoint terminal conditions are enforced by construction, only the three optimality residuals and the state initial mismatch enter the loss:

\begin{equation}
\label{eq:loss_indirect}
\mathcal{L}_{\text{ind}}(\theta,\psi,\phi) = w_y \mathcal{L}_y + w_\lambda \mathcal{L}_\lambda + w_u \mathcal{L}_u + w_{\text{ic}} \mathcal{L}_{\text{ic}},
\end{equation}

where

\begin{align}
\mathcal{L}_y &:= \frac{1}{N_{\text{int}}} \sum_{j=1}^{N_{\text{int}}} |\mathcal{R}_y(x_j,t_j;\theta,\psi)|^2, \\
\mathcal{L}_\lambda &:= \frac{1}{N_{\text{int}}} \sum_{j=1}^{N_{\text{int}}} |\mathcal{R}_\lambda(x_j,t_j;\theta,\phi)|^2, \\
\mathcal{L}_u &:= \frac{1}{N_{\text{bc}}} \sum_{j=1}^{N_{\text{bc}}} |\mathcal{R}_u(x_j^b,t_j^b;\psi,\phi)|^2, \\
\mathcal{L}_{\text{ic}} &:= \frac{1}{N_{\text{ic}}} \sum_{i=1}^{N_{\text{ic}}} |Y_\theta(x_i,0) - y_0(x_i)|^2.
\end{align}

Each $\mathcal{L}_\bullet$ is a Monte-Carlo (or quasi-Monte-Carlo) quadrature approximation of a continuous $L^2$-norm of the corresponding residual, e.g. $\mathcal{L}_y \approx \|\mathcal{R}_y\|_{L^2(Q)}^2$; this correspondence is what links the \emph{discrete, trainable} loss to the \emph{continuous} PDE residuals used in the estimates below.

\subsection{Standing Assumptions}

The estimates in this section are conditional on the following standing assumptions, which we state explicitly rather than leaving implicit.

\begin{itemize}
\item[(A1)] \textbf{Well-posedness.} The optimality system \eqref{eq:state_res}--\eqref{eq:bc_lambda_res} admits a unique solution $(y^*,u^*,\lambda^*)$ in an appropriate energy class, and the linearization of this system about $(y^*,\lambda^*)$ satisfies an inf-sup (stability) condition with constant $C_{\text{stab}}$.
\item[(A2)] \textbf{Regularity of the Dirichlet control problem.} Dirichlet boundary control of a parabolic equation is, in general, a low-regularity problem: for boundary data $u\in L^2(\Sigma)$ the state $y$ is typically only a \emph{very weak solution} (transposition solution), and the trace $\partial_n\lambda|_\Sigma$ appearing in \eqref{eq:stat_res} need not exist in the classical sense~\cite{lasiecka2000control,troltzsch2010optimal}. We assume throughout that the data $(y_0, y_d, u^*)$ are smooth enough, and $\Omega$ is regular enough, that $(y^*,\lambda^*)$ lie in $H^{2,1}(Q)$-type spaces in which all traces below are classically well defined. This is a genuine restriction, not a technicality; results that require only $u \in L^2(\Sigma)$ are outside the scope of the estimates below.
\item[(A3)] \textbf{Nonlinearity.} $f\in C^1(\mathbb{R})$ and $f'$ is bounded below, $f'(s)\geq -c_f$ for some $c_f\geq 0$ (satisfied by the Allen--Cahn nonlinearity $f(y)=y^3-y$, for which $f'(s) = 3s^2-1 \geq -1$).
\item[(A4)] \textbf{Approximation capacity.} The network architectures for $Y_\theta, U_\psi, \Lambda_\phi$ are rich enough that the best-approximation error $\epsilon_{\text{net}}$ defined in \eqref{eq:net_approx} can be made arbitrarily small by increasing width/depth. No explicit rate in network size is claimed here; this is treated as an assumption, not a proven property of the architecture.
\end{itemize}

\subsection{Decomposition of the Total Error}

Let $e_y := y^* - Y_\theta$, $e_\lambda := \lambda^* - \Lambda_\phi$, $e_u := u^* - U_\psi$ denote the errors in the state, adjoint, and control, relative to the exact solution of the optimality system (not to a discretized/mesh solution). The total error is measured in the norm

\begin{equation}
\| (e_y, e_\lambda, e_u) \|_{\mathcal{E}}^2 := \| e_y \|_{L^2(0,T;H^1(\Omega))}^2 + \| e_\lambda \|_{L^2(0,T;H^1(\Omega))}^2 + \| e_u \|_{L^2(\Sigma)}^2.
\end{equation}

Using the triangle inequality and the stability assumption (A1), the total error can be decomposed into three main contributions:

\begin{equation}
\label{eq:decomp}
\| (e_y, e_\lambda, e_u) \|_{\mathcal{E}} \lesssim \mathcal{E}_{\text{approx}} + \mathcal{E}_{\text{opt}} + \mathcal{E}_{\text{quad}},
\end{equation}

where:
\begin{itemize}
    \item $\mathcal{E}_{\text{approx}}$: best-approximation error of the neural network architecture (item (A4)),
    \item $\mathcal{E}_{\text{opt}}$: optimization error, i.e.\ how far the trained loss \eqref{eq:loss_indirect} is from its infimum over the chosen architecture,
    \item $\mathcal{E}_{\text{quad}}$: quadrature (sampling) error between the Monte-Carlo loss $\mathcal{L}_{\text{ind}}$ and the continuous residual norms it estimates.
\end{itemize}

We do not list a separate boundary or terminal contribution because these conditions are hard-constrained. The state initial mismatch is retained explicitly in the trainable loss and in the estimator below.

\subsection{A Priori Error Estimate}

\begin{theorem}[Conditional a priori estimate]
\label{thm:apriori}
Under Assumptions (A1)--(A3), let $(y^*, u^*, \lambda^*)$ be the exact solution of \eqref{eq:state_res}--\eqref{eq:bc_lambda_res}. Suppose the networks $Y_\theta$, $U_\psi$, $\Lambda_\phi$ can approximate $(y^*,\lambda^*,u^*)$ to accuracy $\epsilon_{\text{net}}$ in the $\mathcal{E}$-norm,
\begin{equation}
\label{eq:net_approx}
\inf_{\theta,\psi,\phi} \| (y^* - Y_\theta, \lambda^* - \Lambda_\phi, u^* - U_\psi) \|_{\mathcal{E}} \leq \epsilon_{\text{net}},
\end{equation}
and suppose training reaches parameters $(\theta^*,\psi^*,\phi^*)$ with continuous residual norm
\begin{equation}
\|\mathcal{R}_y\|_{L^2(Q)}^2 + \|\mathcal{R}_\lambda\|_{L^2(Q)}^2 + \|\mathcal{R}_u\|_{L^2(\Sigma)}^2 \;\leq\; \epsilon_{\text{loss}}^2.
\end{equation}
Then
\begin{equation}
\label{eq:apriori_estimate}
\| (y^* - Y_\theta^*, \lambda^* - \Lambda_\phi^*, u^* - U_\psi^*) \|_{\mathcal{E}} \;\lesssim\; \epsilon_{\text{net}} + C_{\text{stab}}\,\epsilon_{\text{loss}},
\end{equation}
with $C_{\text{stab}}$ the stability constant of Assumption (A1).
\end{theorem}

\begin{proof}[Sketch of proof]
Write $e=(e_y,e_\lambda,e_u)$ for the error at the trained parameters, and split $e = e^{\mathrm{approx}} + e^{\mathrm{res}}$, where $e^{\mathrm{approx}}$ is the error of the \emph{best} network approximation (controlled by $\epsilon_{\text{net}}$ via (A4)) and $e^{\mathrm{res}} := (y^*-Y_\theta^*,\dots) - e^{\mathrm{approx}}$ is what remains. By linearity, $e^{\mathrm{res}}$ solves the linearization of \eqref{eq:state_res}--\eqref{eq:stat_res} about $(y^*,\lambda^*)$ with right-hand side equal to $(\mathcal{R}_y,\mathcal{R}_\lambda,\mathcal{R}_u)$ evaluated at the trained networks (this uses (A2) to make the linearization and its traces meaningful, and (A3) to control the sign of the reaction term $f'(y^*)$ so that the linearized adjoint operator is coercive up to a lower-order term, via a standard exponential-weight / G\r{a}rding-inequality argument). The stability assumption (A1) then gives $\|e^{\mathrm{res}}\|_{\mathcal{E}} \leq C_{\text{stab}}\big(\|\mathcal{R}_y\|_{L^2(Q)} + \|\mathcal{R}_\lambda\|_{L^2(Q)} + \|\mathcal{R}_u\|_{L^2(\Sigma)}\big) \leq C_{\text{stab}}\,\epsilon_{\text{loss}}$. Combining with $\|e^{\mathrm{approx}}\|_{\mathcal{E}}\leq \epsilon_{\text{net}}$ and the triangle inequality gives \eqref{eq:apriori_estimate}. We emphasize that this is a \emph{sketch}: a full proof requires making (A1)--(A2) precise (function spaces, trace theorems) for the specific $f$ and domain considered, which is not carried out here.
\end{proof}

\begin{remark}
Bound \eqref{eq:apriori_estimate} uses the \emph{continuous} residual norm $\epsilon_{\text{loss}}$, not the Monte-Carlo loss $\mathcal{L}_{\text{ind}}$ that is actually minimized during training. The gap between the two is exactly $\mathcal{E}_{\text{quad}}$ in \eqref{eq:decomp}; see Section~3.6 for how it is controlled in practice.
\end{remark}

\begin{remark}
For the Allen--Cahn nonlinearity $f(y)=y^3-y$, $f'(s)=3s^2-1\geq-1$ is bounded below, which is what makes the coercivity argument in the proof sketch available; a genuine coercivity estimate for the linearized adjoint operator additionally requires either a smallness condition on $T$ or an exponential time-weight $e^{-c_f t}$ to absorb the $-1$ reaction term, which we do not carry out in detail here.
\end{remark}

\subsection{A Posteriori Error Estimation}

Because $Y_\theta, U_\psi, \Lambda_\phi$ are smooth functions on all of $\overline{Q}$ (for smooth activation functions), there is no computational mesh and no notion of an inter-element jump in the normal derivative — such jumps are identically zero for a globally differentiable network. We therefore build the a posteriori indicator directly from pointwise residual evaluation on a \emph{verification point set}, distinct from (and typically much larger/denser than) the training collocation set, rather than from a finite-element-style mesh.

\subsubsection{Verification-Point Residual Indicator}

Let $\widehat{Q} = \{(x_k,t_k)\}_{k=1}^{K}\subset Q$ and $\widehat{\Sigma} = \{(x_k^b,t_k^b)\}_{k=1}^{K_b}\subset \Sigma$ be dense verification point sets (e.g.\ a fixed low-discrepancy or grid sample, held out from training) with associated quadrature weights $\{\omega_k\}$, $\{\omega_k^b\}$ summing to $|Q|$ and $|\Sigma|$ respectively. Define the global a posteriori indicator

\begin{equation}
\label{eq:estimator_global}
\eta^2 \;:=\; \sum_{k=1}^{K} \omega_k\,|\mathcal{R}_y(x_k,t_k)|^2
\;+\; \sum_{k=1}^{K} \omega_k\,|\mathcal{R}_\lambda(x_k,t_k)|^2
\;+\; \sum_{k=1}^{K_b} \omega_k^b\,|\mathcal{R}_u(x_k^b,t_k^b)|^2
\;+\; \eta_{\text{ic}}^2,
\end{equation}

where $\eta_{\text{ic}}^2$ approximates $\|Y_\theta(\cdot,0)-y_0\|_{L^2(\Omega)}^2$. The adjoint terminal mismatch is identically zero by \eqref{eq:hard_adjoint}. Thus $\eta$ is evaluated on points independent of training and approximates the continuous residual quantity entering Theorem~\ref{thm:apriori}. A post-hoc partition may be used only to visualize local contributions; it introduces no jump terms.

\subsubsection{Reliability}

\begin{theorem}[Reliability]
\label{thm:reliability}
Under Assumptions (A1)--(A3), and with $\eta$ as in \eqref{eq:estimator_global} computed on a sufficiently fine verification set,
\begin{equation}
\| (e_y, e_\lambda, e_u) \|_{\mathcal{E}} \;\lesssim\; \epsilon_{\rm net}+C_{\rm stab}\bigl(\eta + \mathcal{E}_{\text{quad}}\bigr),
\end{equation}
where $\mathcal{E}_{\text{quad}}$ is the (typically negligible, if $K,K_b$ are large) quadrature error of \eqref{eq:estimator_global} itself relative to the exact $L^2$ residual norms.
\end{theorem}

\begin{proof}[Sketch of proof]
This follows conditionally from Theorem~\ref{thm:apriori} by identifying $\eta$ as a quadrature approximation of the continuous residual norm and retaining both the best-approximation and verification-quadrature contributions.
\end{proof}

\subsubsection{On "Efficiency" (Lower Bound)}

A finite-element a posteriori estimator is typically also \emph{efficient}: $\eta(E) \lesssim \|e\|_{1,D_E}$ up to data oscillation, proved with element-wise bubble functions supported on individual mesh cells. This construction relies essentially on having a mesh with compactly supported, element-local test functions. In the present meshfree PINN setting there is no such local test-function machinery available in general, and we are not aware of a proof of a matching lower bound for collocation-based PINN residual indicators of this type. We therefore do \emph{not} claim an efficiency theorem here; \eqref{eq:estimator_global} should be understood as a reliable (upper-bounding, up to constants) indicator only, useful for monitoring convergence and flagging poorly-resolved regions of $Q$, but not as a certified two-sided estimator. Establishing efficiency rigorously in the meshfree setting is left as an open direction.

\subsection{Loss Function as an Error Indicator}

Combining Theorem~\ref{thm:apriori} with the identification of $\eta$ as a quadrature estimate of $\epsilon_{\text{loss}}$, we have informally
\begin{equation}
\| (e_y, e_\lambda, e_u) \|_{\mathcal{E}}^2 \;\lesssim\; \mathcal{L}_{\text{ind}} + \mathcal{E}_{\text{approx}} + \mathcal{E}_{\text{quad}},
\end{equation}
so a small training loss is \emph{necessary but not sufficient} for a small true error: it must be accompanied by (i) a sufficiently expressive architecture ($\mathcal{E}_{\text{approx}}$ small) and (ii) a sufficiently dense/well-distributed collocation set ($\mathcal{E}_{\text{quad}}$ small). In particular, a low training loss evaluated \emph{on the training points themselves} can be misleading if the network overfits those points; this is precisely why $\eta$ in Section~3.5 is evaluated on an independent verification set.

\subsection{Practical Implementation of the Error Estimator}

The a posteriori indicator is computed after training as follows:
\begin{enumerate}
    \item Sample a verification point set $\widehat{Q}\cup\widehat{\Sigma}$, independent of the training collocation points (e.g.\ a fresh random draw, or a fixed low-discrepancy grid), together with quadrature weights.
    \item Evaluate $Y_\theta, U_\psi, \Lambda_\phi$ and their required derivatives at these points via automatic differentiation, and compute the residuals $\mathcal{R}_y, \mathcal{R}_\lambda, \mathcal{R}_u$.
    \item Form the weighted sum \eqref{eq:estimator_global}.
    \item (Optional, diagnostic only) Bin the pointwise squared residuals over a post-hoc space-time partition purely for visualization of where the error is concentrated; this partition does not enter the estimator's value.
\end{enumerate}
The computational cost is comparable to one evaluation of the loss function on $K+K_b$ points, i.e.\ negligible relative to training.

\subsection{Numerical Validation Strategy}

To validate the error indicator, we propose the following procedure:
\begin{enumerate}
    \item Construct a reference solution $(y_{\text{ref}}, u_{\text{ref}}, \lambda_{\text{ref}})$ with a fine-grid FEM/adjoint solver.
    \item Train the Indirect PINN with an increasing number of training collocation points $N = N_{\text{int}} = N_{\text{bc}}$ (there is no mesh size $h$ in the meshfree setting; $N$ plays the analogous role).
    \item Compute the true error $\|e\|_{\mathcal{E}}$ against the reference solution.
    \item Compute the indicator $\eta$ from \eqref{eq:estimator_global} on a verification set independent of both the $N$ training points and the reference-solution grid.
    \item Compare $\eta$ with the true error as $N$ increases, and report the empirical ratio $\eta/\|e\|_{\mathcal{E}}$ (an \emph{effectivity index}) rather than assuming a fixed algebraic rate; unlike FEM, PINN convergence in $N$ is not generally guaranteed to follow a clean power law, and any observed rate should be reported empirically rather than asserted a priori.
\end{enumerate}
This validates the reliability of the estimator in the sense of Theorem~\ref{thm:reliability}; it does not, and is not intended to, validate an efficiency claim (see Section~3.5.3).

\subsection{Summary of the Error Estimation Framework}

The proposed error estimation framework provides:
\begin{itemize}
    \item \textbf{A conditional a priori bound} (Theorem~\ref{thm:apriori}), valid under explicit well-posedness, regularity, and approximation-capacity assumptions (A1)--(A4).
    \item \textbf{A computable a posteriori indicator} \eqref{eq:estimator_global}, built from pointwise residual evaluation on a verification set, with no artificial mesh-jump terms.
    \item \textbf{A reliability theorem} (upper bound), together with an honest statement that a matching efficiency (lower bound) result is not established here for the meshfree setting.
    \item \textbf{Practical guidelines} for computing and validating the indicator empirically.
\end{itemize}

\begin{remark}
The Dirichlet boundary control setting analyzed here (Assumption (A2)) is a known source of low-regularity difficulties for the classical PDE theory; results here should not be read as removing that difficulty, only as working under regularity hypotheses strong enough to bypass it. Extending the framework to genuinely low-regularity Dirichlet data, and to a provably efficient meshfree a posteriori indicator, are left for future work.
\end{remark}
